Natural selection in contemporary humans is linked to income and substitution effects

### 2021-01-11 why-natural-selection.Rmd
Tweaks for UKBB; formatting.

---
title: "Natural selection in contemporary humans is linked to income and substitution effects"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against). Selection effects are concentrated among lower-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects and that are strongly related to the socio-economic system."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

These patterns are in line with economic theories of fertility
[@becker1960economic]. In these, higher potential earnings have two opposite
effects on fertility: a fertility-increasing *income effect* (higher income
makes children more affordable), and a fertility-lowering *substitution effect*
(time spent on childrearing has a higher cost in foregone earnings). Our
results suggest that the substitution effect dominates for single parents,
while the income effect is stronger for couples.

@@

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center"}

sexes_theme <-  my_theme(
                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
                  legend.position      = "bottom", 
                  legend.box.spacing   = unit(0, "lines"),
                  legend.box.margin    = margin(-5, 0, 0, -5),
                  legend.margin        = margin(0, 0, 0, 0),
                  legend.box.just      = "left",
                  legend.justification = "left"
                )
sexes_guides <- guides(
                  # fill  = guide_legend(title.position = "top"),
                  # color = guide_legend(title.position = "top")
                )

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their effect
on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each polygenic score against that score's correlation with earnings in a respondent's first
job, and against its correlation with educational attainment, a
predictor of lifetime earnings. The raw relationships (left column) are strongly
negative. If we plot selection effects controlling for age at first live birth
(right column), the effect reverses and becomes positive. Thus, the labour
market appears to play an important role in natural selection. Substitution
effects dominate income effects overall, which fits the known association
between income and lower fertility [@becker1960economic;@jones2006economic].

@@

Previous work has documented natural selection in modern populations on variants
underlying polygenic traits [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including people with lower income, lower education, younger first parenthood, 
and more lifetime sexual partners. Indeed, among older mothers, selection effects are
actually reversed. Furthermore, the size of selection effects on a polygenic
score correlates with that score's association with labour market earnings. The
economic theory of fertility provides a parsimonious explanation for this.

@@

Because selection effects are concentrated in poorer groups, they may also increase
inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with 
educational attainment, which lowers the mean, is stronger at the bottom of the 
income distribution, and this increases the differences between groups.  Overall,
natural selection increases inequality for 29 out of 33 polygenic scores 
(Appendix Table \@ref(tab:tbl-inequality-effects)).

>>>>

---
title: "Natural selection in contemporary humans is linked to income and substitution effects"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against). Selection effects are concentrated among lower-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

These patterns are in line with economic theories of fertility
[@becker1960economic]. In these, higher potential earnings have two opposite
effects on fertility: a fertility-increasing *income effect* (higher income
makes children more affordable), and a fertility-lowering *substitution effect*
(time spent on childrearing has a higher cost in foregone earnings). Our
results suggest that the substitution effect dominates for single parents and
younger parents, while among couples and older parents (22+) effects are more
evenly balanced.

@@

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center", fig.height = 4}

sexes_theme <-  my_theme(
                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
                  legend.position      = "bottom", 
                  legend.box.spacing   = unit(0, "lines"),
                  legend.box.margin    = margin(-5, 0, 0, -85),
                  legend.margin        = margin(0, 0, 0, 0),
                  legend.box.just      = "left",
                  legend.justification = "left",
                  legend.spacing       = unit(1.5, "lines")
                )
sexes_guides <- guides(
                  # fill  = guide_legend(ncol = 1),
                  # color = guide_legend(ncol = 1)
                )

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their effect
on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each polygenic score against that score's correlation with earnings in a respondent's first
job, and against its correlation with educational attainment, a
predictor of lifetime earnings.^[animation] The raw relationships (left column) are strongly
negative. If we plot selection effects controlling for age at first live birth
(right column), the effect reverses and becomes positive. Thus, the labour
market appears to play an important role in natural selection. Substitution
effects dominate income effects overall, which fits the known association
between income and lower fertility [@becker1960economic;@jones2006economic].

^[animation]: See also the supplementary animation at https://xxx.xxx.xxx

@@

Previous work has documented natural selection in modern populations on variants
underlying polygenic traits [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including people with lower income, lower education, younger first parenthood, 
and more lifetime sexual partners. Indeed, among older mothers (22+), selection effects are
actually reversed. Furthermore, the size of selection effects on a polygenic
score correlates with that score's association with labour market earnings. The
economic theory of fertility provides a parsimonious explanation for this.

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups.
Overall, natural selection increases inequality for 29 out of 33 polygenic
scores (Appendix Table \@ref(tab:tbl-inequality-effects)). Since many polygenic
scores are predictive of disease risk, this increase in inequality could
potentially increase health inequalities. In general, the demographic history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may help in the design
of health policies.
